hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-06-18 inst/manuscript/paper.Rmd
output figures to separat file,  not inline
add figures and tables at end of manuscript, no longer inline

@@

```{r, echo = FALSE, out.height= "72%", fig.align = 'center', fig.cap="hatchR workflow. Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decision choices as italicized text."}
knitr::include_graphics("./workflow.png")
```

@@

```{r, echo = FALSE, fig.height=3, fig.align = 'center', fig.cap = "Effective value model fit and application for European Grayling. Panel A shows developmental rates (days to hatch) at five differrent average rearing temperatures and the corresponding model fit (blue line). Panel B uses the effective value model derived from the blue line in panel A to predict the effective value of daily average temperatures ranging from 2 to 14 °C. The dashed line is a 0.01 increase in effective value for every 1 °C increase as a point of reference."}
grayling <- matrix(NA, 5, 2) |> tibble::tibble()
colnames(grayling) <- c("temp", "days")

@@

p1 + p2 + plot_annotation(tag_levels = "A")
```

@@

knitr::kable(df,
  booktabs = TRUE,
  caption = "Example temperature data for use in hatchR."
)
```

@@

```{r, echo=TRUE}
# apply predict_phenology() over multiple sites and spawn dates
hatch_res <- isaak_summ_bt |>
  mutate(
    dev_period = map2(
      summ_obj,                  # summarized temperature data object
      spawn_dates,               # list of spawn dates
      predict_phenology,         # function to apply
      temperature = daily_temp,  # column of temperature data in `summ_obj`
      dates = date,              # column of date in `summ_obj`
      model = bt_hatch           # model to use (from model_select())
    ) |>
      map_df("dev_period") |>    # pull out just dev_period results
      list()
  ) |>
  select(site, dev_period) |>           # just select the columns we want
  unnest(cols = c(dev_period)) |>       # un-nest everything
  mutate(days_to_hatch = stop - start)  # make a new column of days to hatch
```

@@

```{r plot-bt, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap="Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data."}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

@@

ggplot(
  data = hatch_factors,
  aes(
    x = days_to_hatch,
    y = spawn_time,
    fill = spawn_time,
    color = spawn_time
  )
) +
  geom_density_ridges(alpha = 0.9) +
  facet_wrap(~year, ncol = 1) +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  scale_color_brewer(palette = "Blues", direction = 1) +
  labs(x = "Days to hatch", y = "Spawn time") +
  theme_classic() +
  theme(legend.position = "none")
```

@@

```{r plot-check-1, echo = TRUE,  fig.cap = "Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset. Users can set custom thresholds for minimum and maximum temperatures (dashed lines).", fig.height = 3.25}
# visual check of temperature data
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0,
                temp_max = 25)
```

@@

```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.", fig.height=5}
# Catfish
catfish_res <- predict_phenology(
  data = data_sim,
  dates = date_sim,
  temperature = temp_sim,
  spawn.date = "2000-07-01",
  model = cat_mod$expression
)

@@

ggplot() +
  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim)) +
  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim)) +
  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax = index - .35, ymin = index - .5, fill = Species)) +
  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index - 0.425), label = days)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
  theme_classic() +
  theme(legend.position = c(0.75, 0.25))
```

@@

The **hatchR software** includes a built in function, `plot_phenology()`, for visualizing phenology predictions. This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span. By default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles. However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results. This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.

```{r, fig.cap="Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset."}
plot_phenology(WI_hatch)

@@

@@

>>>>

## Workflow

@@

```{r, echo = FALSE, eval=FALSE, out.height= "72%", fig.align = 'center', fig.cap=""}
knitr::include_graphics("./workflow.png")
```

@@

```{r, echo = FALSE, fig.height=3, fig.align = 'center', fig.cap = ""}
grayling <- matrix(NA, 5, 2) |> tibble::tibble()
colnames(grayling) <- c("temp", "days")

@@

figure_2 <- p1 + p2 + plot_annotation(tag_levels = "A")
ggsave("./manuscript_figures/Figure_2.pdf", figure_2, width = 8, height = 4, dpi = 1000)
```

@@

@@

```{r, echo=FALSE}
# apply predict_phenology() over multiple sites and spawn dates
hatch_res <- isaak_summ_bt |>
  mutate(
    dev_period = map2(
      summ_obj,                  # summarized temperature data object
      spawn_dates,               # list of spawn dates
      predict_phenology,         # function to apply
      temperature = daily_temp,  # column of temperature data in `summ_obj`
      dates = date,              # column of date in `summ_obj`
      model = bt_hatch           # model to use (from model_select())
    ) |>
      map_df("dev_period") |>    # pull out just dev_period results
      list()
  ) |>
  select(site, dev_period) |>           # just select the columns we want
  unnest(cols = c(dev_period)) |>       # un-nest everything
  mutate(days_to_hatch = stop - start)  # make a new column of days to hatch
```

@@

```{r plot-bt, message=FALSE, echo=FALSE,  out.width='75%', fig.align='center', fig.cap=""}
cut_ints <- date(
  c(
    "2011-08-31", "2012-08-31",
    "2013-08-31", "2014-08-31",
    "2015-08-31"
  )
)

@@

figure_3 <- ggplot(
  data = hatch_factors,
  aes(
    x = days_to_hatch,
    y = spawn_time,
    fill = spawn_time,
    color = spawn_time
  )
) +
  geom_density_ridges(alpha = 0.9) +
  facet_wrap(~year, ncol = 1) +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  scale_color_brewer(palette = "Blues", direction = 1) +
  labs(x = "Days to hatch", y = "Spawn time") +
  theme_classic() +
  theme(legend.position = "none")

ggsave("./manuscript_figures/Figure_3.pdf", figure_3, width = 3.5, height = 5, dpi = 1000)
```

@@

```{r plot-check-1, echo = TRUE, eval=FALSE, fig.cap = "", fig.height = 3.25}
# visual check of temperature data
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0,
                temp_max = 25)

```{r, echo=FALSE}
box_1 <- plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0,
                temp_max = 25)

ggsave("./manuscript_figures/Box_1.pdf", box_1, width = 6, height = 3, dpi = 1000)
```

@@

```{r pred-pheno-warmwater, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="", fig.height=5}
# Catfish
catfish_res <- predict_phenology(
  data = data_sim,
  dates = date_sim,
  temperature = temp_sim,
  spawn.date = "2000-07-01",
  model = cat_mod$expression
)

@@

box_4 <- ggplot() +
  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim)) +
  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim)) +
  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax = index - .35, ymin = index - .5, fill = Species)) +
  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index - 0.425), label = days)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
  theme_classic() +
  theme(legend.position = c(0.75, 0.25))

ggsave("./manuscript_figures/Box_4.pdf", box_4, width = 4, height = 4, dpi = 1000)
```

@@

The **hatchR software** includes a built in function, `plot_phenology()`, for visualizing phenology predictions from either `predict_phenology()` or `predict_spawn()`. This function generates plots with three specific components: 1) the temperature regime over the prediction period, 2) the cumulative sum of effective values, and 3) the effective value for each day within the prediction span. By default, `plot_phenology()` produces a comprehensive figure that includes all three elements with corresponding labels and titles. However, users can customize the output to focus on specific aspects of interest, allowing for tailored visual representations of their results. This function provides a quick and effective way to interpret model outputs, facilitating comparisons between temperature regimes or species-specific phenological responses.

```{r, eval = FALSE, fig.cap=""}
plot_phenology(WI_hatch)

@@

```{r, echo=FALSE}
box_5 <- plot_phenology(WI_hatch)
ggsave("./manuscript_figures/Box_5.pdf", box_5, width = 6, height = 6, dpi = 1000)
```

@@

::: {#refs}
:::

# Table Captions

**Table 1.** Example temperature data for use in hatchR.

# Figure Captions

**Figure 1.** hatchR workflow. Data inputs are represented by folders, data processes by filled circles, hatchR functions as plain text rectangles, and decision choices as italicized text.

**Figure 2.** Effective value model fit and application for European Grayling. Panel A shows developmental rates (days to hatch) at five differrent average rearing temperatures and the corresponding model fit (blue line). Panel B uses the effective value model derived from the blue line in panel A to predict the effective value of daily average temperatures ranging from 2 to 14 °C. The dashed line is a 0.01 increase in effective value for every 1 °C increase as a point of reference.

**Figure 3.** Predicted days to hatch for 139 putative bull trout populations over three spawning periods (Early = September 1, Peak = September 15, Late = September 30) and four years of temperature data.

**Box 1.** Output of hatchR function `plot_check_temp()`, which is used as a visual data check on the raw `year_sim` dataset. Users can set custom thresholds for minimum and maximum temperatures (dashed lines).

**Box 4.** Predicted days to hatch for three warmwater species with custom parameterizations using a random thermal regime with an ascending thermograph with a mean temperature of 16 °C over 30 days.

**Box 5.** Output of `plot_phenolgy()` function using predicted hatch time from `woody_island` dataset.

# Tables

knitr::kable(df,
  booktabs = TRUE
)
```
